---
output:  
    word_document :
        reference_docx: QaTableStyle6.docx
params: 
    DBServerName: 
        label: "Database Server Name:"
        value: cc-s-d05.ucdenver.pvt
        input: select
        choices: [cc-s-d05.ucdenver.pvt, vwphbisql06, ihrsql1.ihr.or.kp.org]
    DBName:
        label: "Database Name:"
        value: VDW_3_1_DH
        input: select
        choices: [VDW_3_1_DH, VDW.dbo, VDW]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
startTime <- Sys.time()

# Names of database tables
demographics    <- "demographics"
encounters      <- "encounters"
census_location <- "census_location"
diagnoses       <- "diagnoses"
vital_signs     <- "vital_signs"
lab_results     <- "lab_results"
procedures      <- "procedures"

# set to zero for full read of the database
# set to a positive number to limit the number of rows read from each table
maxQryRows <- 0

# load lib for SQL server connectivity
# install.packages("RODBC")
library(RODBC)

library(knitr)
library(tidyverse)

# connection to sql server
con <- DBI::dbConnect(odbc::odbc(),
                      Driver = "SQL Server",
                      Server = params$DBServerName,
                      Database = params$DBName)

# connection to sql server
sqldb <- odbcDriverConnect(paste('driver={SQL Server};server=',params$DBServerName,';database=',params$DBName, sep=""), readOnlyOptimize = TRUE)

# run this to close the connection to the database
# close(sqldb)

```

#CHORDS QA Report: VDW P1 Tables

The purpose of the data quality program is to characterize the data in CHORDS VDW 3.1 priority level 1 (P1) tables. The P1 tables include the following: DEMOGRAPHICS, ENCOUNTERS, CENSUS_LOCATION, DIAGNOSES, and VITAL_SIGNS. The program uses a series of SQL queries operationalized using RStudio to produce this report. These tables provide descriptive information about data stored in a data partner’s VDW and can be used to assess data model conformance, data plausibility, and data completeness. This data quality report was generated from CHORDS `r params$DBName`.

###Information about the QA program 
Data Partner: 
Analyst: 
Query Run Date:  `r Sys.Date()`


```{r supporting data and functions, include = FALSE}
## helper functions
 # returns a vector with the column numbers of character variables in the data frame
charVars <- function(df) grep('^ch',sapply(df,class))

 # trims all character variables in a dataframe, sets blank to NA
trimChrVars <- function(df){
for(i in charVars(df)){
  df[,i] <- gsub('\\s+$','',df[,i])
  df[,i] <- ifelse(nchar(df[,i])==0,NA,df[,i])
}
  df
}

# age category calculator
ageCatCalc <- function(age){
  ageCat <- factor(
    ifelse(is.na(age)==TRUE, 0,
           ifelse(age<0              , 1, 
                  ifelse(age>=0 & age<2, 2,
                         ifelse(age>=2 & age<5, 3, 
                                ifelse(age>=5 & age<10, 4, 
                                       ifelse (age>=10 & age<15, 5, 
                                               ifelse(age>=15 & age<19, 6,
                                                      ifelse(age>=19 & age<22, 7,
                                                             ifelse(age>=22 & age<45, 8, 
                                                                    ifelse(age>=45 & age<65, 9, 
                                                                           ifelse(age>=65 & age<75, 10, 
                                                                                  ifelse(age>=75 & age<90, 11,
                                                                                         ifelse(age>=90 , 12, 13)))))) ))))))),
    levels=0:13,
    labels=c('Missing','Negative','0-1','2-4','5-9','10-14','15-18','19-21','22-44','45-64','65-74','75-89','90+','Other')
  )
  
  return(ageCat)
}

# supporting look up tables

stateCnty <- chordsTables::stateCnty

ccsMDxI9 <- chordsTables::ccsMDxI9

ccsMDxI10 <- chordsTables::ccsMDxI10

ccsSDxI9 <- chordsTables::ccsSDxI9

icdLU <- rbind(ccsMDxI10[c('dxCdTp','dxDesc')],ccsSDxI9[c('dxCdTp','dxDesc')])

# ISO language lookup
isoLang <- chordsTables::isoLang

# VDW value sets
valSets <- chordsTables::valSets


```

## Data Quality Report Results

##Table 1. Records, Patients, Encounters, and Date Ranges by Table

This table contains summary counts and date ranges by table. Distinct encounters and patient are shown for applicable tables. Data ranges should be used to compare data time windows between tables.

```{r tab1_sum_stats}

tab1_dem <- odbc::dbGetQuery(con, "select count(*) as nrows, 
	count(distinct person_id) as npats,
  'RESULT_DT' as fieldname,
	min(result_dt) as mindt,
	max(result_dt) as maxdt
	from demographics")

tab1_dem <- data.frame(nrows = tab1_lab$nrows, 
                       npats = tab1_lab$npats, 
                       nencts = NA, 
                       fieldname = tab1_lab$fieldname, 
                       mindt = tab1_lab$mindt, 
                       maxdt = tab1_lab$maxdt)


tab1_enc <- odbc::dbGetQuery(con, "select count(*) as nrows, 
	count(distinct person_id) as npats,
  count(distinct enc_id) as nencts,
  'RX_ORDER_DATE' as fieldname,
	min(rx_order_date) as mindt,
	max(rx_order_date) as maxdt
	from encounters")


tab1_px <- odbc::dbGetQuery(con, "select count(*) as nrows, 
	count(distinct person_id) as npats,
  count(distinct enc_id) as nencts,
  'PROCDATE' as fieldname,
	min(procdate) as mindt,
	max(procdate) as maxdt
	from procedures")


tab1_sh <- odbc::dbGetQuery(con, "select count(*) as nrows, 
	count(distinct person_id) as npats,
  'CONTACT_DATE' as fieldname,
  count(distinct enc_id) as nencts,
	min(contact_date) as mindt,
	max(contact_date) as maxdt
	from social_history")

tab1_ps <- odbc::dbGetQuery(con, "select count(*) as nrows
	from provider_specialty")

tab1_ps <- data.frame(nrows = tab1_ps$nrows, npats = NA, nencts = NA, fieldname = NA, mindt = NA, maxdt = NA)

tab1_all <- bind_rows(tab1_lab, tab1_rx, tab1_px, tab1_sh, tab1_ps) %>% 
  bind_cols(Table = c("LAB_RESULTS", "PRESCRIBING", "PROCEDURES", "SOCIAL_HISTORY", "PROVIDER_SPECIALTY"), .)

```

```{r tab1_disp, results="asis"}

kable(tab1_all, col.names = c("Table", "Records", "Patients", "Encounters", "Date Field", "Min Date", "Max Date"), format.args = list(big.mark = ",")) 
  
```

##Table 2: Missingness Variables across P1 Tables

This table contains record counts for null and unknown values across P2 tables.


```{r tab2_res}
tab2_lab_null <- odbc::dbGetQuery(con, 
  "SELECT SUM(CASE WHEN TEST_TYPE IS NULL THEN 1 ELSE 0 END) as TEST_TYPE,
   SUM(CASE WHEN BATTERY_CD IS NULL THEN 1 ELSE 0 END) as BATTERY_CD,
   SUM(CASE WHEN LOINC IS NULL THEN 1 ELSE 0 END) as LOINC,
   SUM(CASE WHEN LOCAL_CD IS NULL THEN 1 ELSE 0 END) as LOCAL_CD,
   SUM(CASE WHEN RESULT_NUM IS NULL THEN 1 ELSE 0 END) as RESULT_NUM,
   SUM(CASE WHEN RESULT_UNIT IS NULL THEN 1 ELSE 0 END) as RESULT_UNIT,
   SUM(CASE WHEN SPECIMEN_SOURCE IS NULL THEN 1 ELSE 0 END) as SPECIMEN_SOURCE
      FROM lab_results")

tab2_lab_null2 <- data.frame(table = rep("LAB_RESULTS", ncol(tab2_lab_null)), var = names(tab2_lab_null), records_null = t(tab2_lab_null)[,1])
row.names(tab2_lab_null2) <- NULL

tab2_rx_null <- odbc::dbGetQuery(con, 
  "SELECT SUM(CASE WHEN GENERIC_MED_NAME IS NULL THEN 1 ELSE 0 END) as GENERIC_MED_NAME,
   SUM(CASE WHEN RX_DAYS_SUPPLY IS NULL THEN 1 ELSE 0 END) as RX_DAYS_SUPPLY,
   SUM(CASE WHEN RX_END_DATE IS NULL THEN 1 ELSE 0 END) as RX_END_DATE,
   SUM(CASE WHEN RX_START_DATE IS NULL THEN 1 ELSE 0 END) as RX_START_DATE,
   SUM(CASE WHEN RX_QUANTITY IS NULL THEN 1 ELSE 0 END) as RX_QUANTITY,
   SUM(CASE WHEN RX_QUANTITY_NUM IS NULL THEN 1 ELSE 0 END) as RX_QUANTITY_NUM,
   SUM(CASE WHEN RX_QUANTITY_UNIT IS NULL THEN 1 ELSE 0 END) as RX_QUANTITY_UNIT,
   SUM(CASE WHEN RX_REFILLS IS NULL THEN 1 ELSE 0 END) as RX_REFILLS,
   SUM(CASE WHEN RXMD IS NULL THEN 1 ELSE 0 END) as RXMD,
   SUM(CASE WHEN RXNORM IS NULL THEN 1 ELSE 0 END) as RXNORM
      FROM prescribing")

tab2_rx_null2 <- data.frame(table = rep("PRESCRIBING", ncol(tab2_rx_null)), var = names(tab2_rx_null), records_null = t(tab2_rx_null)[,1])
row.names(tab2_rx_null2) <- NULL

tab2_px_null <- odbc::dbGetQuery(con, 
  "SELECT SUM(CASE WHEN PROVIDER IS NULL THEN 1 ELSE 0 END) as PROVIDER,
   SUM(CASE WHEN ENCTYPE IS NULL THEN 1 ELSE 0 END) as ENCTYPE,
   SUM(CASE WHEN PXCNT IS NULL THEN 1 ELSE 0 END) as PXCNT
      FROM procedures")

tab2_px_null2 <- data.frame(table = rep("PROCEDURES", ncol(tab2_px_null)), var = names(tab2_px_null), records_null = t(tab2_px_null)[,1])
row.names(tab2_px_null2) <- NULL

tab2_ps_null <- odbc::dbGetQuery(con, 
  "SELECT SUM(CASE WHEN SPECIALTY IS NULL THEN 1 ELSE 0 END) as SPECIALTY,
   SUM(CASE WHEN PROVIDER_TYPE IS NULL THEN 1 ELSE 0 END) as PROVIDER_TYPE,
   SUM(CASE WHEN PROVIDER_BIRTH_YEAR IS NULL THEN 1 ELSE 0 END) as PROVIDER_BIRTH_YEAR,
   SUM(CASE WHEN PROVIDER_GENDER IS NULL THEN 1 ELSE 0 END) as PROVIDER_GENDER,
   SUM(CASE WHEN PROVIDER_RACE IS NULL THEN 1 ELSE 0 END) as PROVIDER_RACE,
   SUM(CASE WHEN PROVIDER_HISPANIC IS NULL THEN 1 ELSE 0 END) as PROVIDER_HISPANIC
      FROM provider_specialty")

tab2_ps_null2 <- data.frame(table = rep("PROVIDER_SPECIALTY", ncol(tab2_ps_null)), var = names(tab2_ps_null), records_null = t(tab2_ps_null)[,1])
row.names(tab2_ps_null2) <- NULL

tab2_sh_null <- odbc::dbGetQuery(con, 
  "SELECT SUM(CASE WHEN TOBACCO_USER IS NULL THEN 1 ELSE 0 END) as TOBACCO_USER,
   SUM(CASE WHEN ALCOHOL_USER IS NULL THEN 1 ELSE 0 END) as ALCOHOL_USER,
   SUM(CASE WHEN ILL_DRUG_USER IS NULL THEN 1 ELSE 0 END) as ILL_DRUG_USER,
   SUM(CASE WHEN ENC_ID IS NULL THEN 1 ELSE 0 END) as ENC_ID
      FROM social_history")

tab2_sh_null2 <- data.frame(table = rep("SOCIAL_HISTORY", ncol(tab2_sh_null)), var = names(tab2_sh_null), records_null = t(tab2_sh_null)[,1])
row.names(tab2_sh_null2) <- NULL

tab2_null <- bind_rows(tab2_lab_null2, tab2_rx_null2, tab2_px_null2, tab2_ps_null2, tab2_sh_null2) %>% 
  mutate(
    pct_rec = ifelse(table == "LAB_RESULTS", (records_null / tab1_lab$nrows * 100), 
                     ifelse(table == "PRESCRIBING", (records_null / tab1_rx$nrows * 100), 
                             ifelse(table == "PROCEDURES", (records_null / tab1_px$nrows * 100), 
                                     ifelse(table == "PROVIDER_SPECIALTY", (records_null / tab1_ps$nrows * 100), 
                                             ifelse(table == "SOCIAL_HISTORY", (records_null / tab1_sh$nrows * 100), NA)))))
  )

tab2_lab_uk <- odbc::dbGetQuery(con, 
  "SELECT SUM(CASE WHEN TEST_TYPE = 'Unknown' THEN 1 ELSE 0 END) as TEST_TYPE
      FROM lab_results")

tab2_lab_uk2 <- data.frame(table = rep("LAB_RESULTS", ncol(tab2_lab_uk)), var = names(tab2_lab_uk), records_uk = t(tab2_lab_uk)[,1])
row.names(tab2_lab_uk2) <- NULL


tab2_ps_uk <- odbc::dbGetQuery(con, 
  "SELECT SUM(CASE WHEN SPECIALTY = 'UNK' THEN 1 ELSE 0 END) as SPECIALTY,
   SUM(CASE WHEN PROVIDER_TYPE = 999 THEN 1 ELSE 0 END) as PROVIDER_TYPE,
   SUM(CASE WHEN PROVIDER_GENDER = 'U' THEN 1 ELSE 0 END) as PROVIDER_GENDER,
   SUM(CASE WHEN PROVIDER_RACE = 'U' THEN 1 ELSE 0 END) as PROVIDER_RACE,
   SUM(CASE WHEN PROVIDER_HISPANIC = 'U' THEN 1 ELSE 0 END) as PROVIDER_HISPANIC
      FROM provider_specialty")

tab2_ps_uk2 <- data.frame(table = rep("PROVIDER_SPECIALTY", ncol(tab2_ps_uk)), var = names(tab2_ps_uk), records_uk = t(tab2_ps_uk)[,1])
row.names(tab2_ps_uk2) <- NULL

tab2_sh_uk <- odbc::dbGetQuery(con, 
  "SELECT SUM(CASE WHEN TOBACCO_USER = 'U' THEN 1 ELSE 0 END) as TOBACCO_USER,
   SUM(CASE WHEN ALCOHOL_USER = 'U' THEN 1 ELSE 0 END) as ALCOHOL_USER,
   SUM(CASE WHEN ILL_DRUG_USER = 'U' THEN 1 ELSE 0 END) as ILL_DRUG_USER
      FROM social_history")

tab2_sh_uk2 <- data.frame(table = rep("SOCIAL_HISTORY", ncol(tab2_sh_uk)), var = names(tab2_sh_uk), records_uk = t(tab2_sh_uk)[,1])
row.names(tab2_sh_uk2) <- NULL

tab2_uk <- bind_rows(tab2_lab_uk2, tab2_ps_uk2, tab2_sh_uk2) %>% 
  mutate(
    pct_rec_uk = ifelse(table == "LAB_RESULTS", (records_uk / tab1_lab$nrows * 100), 
                                     ifelse(table == "PROVIDER_SPECIALTY", (records_uk / tab1_ps$nrows * 100), 
                                             ifelse(table == "SOCIAL_HISTORY", (records_uk / tab1_sh$nrows * 100), NA)))
  )

tab2_all <- left_join(tab2_null, tab2_uk, by = c("table", "var"))

```

```{r tab2_disp, results="asis"}

kable(tab2_all, col.names = c("Table", "Variable", "Null Records", "Percent of Records Null", "Unknown Records", "Percent of Records Unknown"), digits = 2, format.args = list(big.mark = ","))
```


#Demographics

##Table 1 . Overall demographics, Dataset: DEMOGRAPHICS

```{r Demographic table, include=F}
nDemogRows <- sqlQuery(sqldb, 'select  count(*) as nDemogRows FROM  demographics') 

demogSummary <- sqlQuery(sqldb, 
                  paste('select  count(*) as nDemogRows ',
                                ', count(distinct person_id) as nPatsDemog ',
		                            ', sum(nPatsWithEncs) as nPatsWithEncs ',
		                            ', sum(nPatsWithLoc) as nPatsWithLoc ',
		                            ', sum(nPatsWithDiag) as nPatsWithDiag ',
		                            ', sum(nPatsWithVital) as nPatsWithVital' ,
		                            '  FROM (select distinct',
		                                          ' person_id',
		                                         ', case when person_id in (select distinct person_id from', encounters,' ) then 1 else 0 end as nPatsWithEncs ',
		                                        ',  case when person_id in (select distinct person_id from', census_location,' ) then 1 else 0 end as nPatsWithLoc ',
		                                        ',  case when person_id in (select distinct person_id from', diagnoses,' ) then 1 else 0 end as nPatsWithDiag ',
		                                        ',  case when person_id in (select distinct person_id from', vital_signs,' ) then 1 else 0 end as nPatsWithVital ',
		                                        '  FROM',  demographics,') as q',
                        sep=" "
                 ),
                 max=maxQryRows
                  )

gc()

demogSummary <- cbind(demogSummary,nDemogRows)

demogTable <-
  with(demogSummary,
  rbind(
    c('Total rows in the demographic table'         ,nDemogRows    ,round(100*nDemogRows/nPatsDemog,1)),
    c('Unique patients in the demographic table'    ,nPatsDemog    ,round(100*nPatsDemog/nPatsDemog,1)),
    c('Unique patients found in CENSUS_LOCATION'    ,nPatsWithLoc  ,round(100*nPatsWithLoc/nPatsDemog,1)),
    c('Unique patients found in DIAGNOSES'          ,nPatsWithDiag ,round(100*nPatsWithDiag/nPatsDemog,1)),
    c('Unique patients found in VITAL_SIGNS'        ,nPatsWithVital,round(100*nPatsWithVital/nPatsDemog,1))
  )
)
```
```{r echo=FALSE}
knitr::kable(demogTable, col.names=c("Characteristic","Frequency","Percent of unique patients"), row.names=FALSE, format.args = list(big.mark = ","))
```

`r ifelse(demogSummary$nDemogRows==demogSummary$nPatsDemog," ","The demographics table has duplicate rows, by person_id")`


## Table 4. 

```{r tab4_res}

tab4 <- odbc::dbGetQuery(con, 
  "select year(rx_order_date) as year,
    count(*) as nrows, 
	  count(distinct person_id) as npats
	  from prescribing
	    group by year(rx_order_date)
      order by year(rx_order_date)")

tab4_avg_pats <- odbc::dbGetQuery(con, 
  "select year(rx_order_date) as year,
    count(*) as nrows
	  from prescribing
	    group by year(rx_order_date), person_id
      order by year(rx_order_date)") %>% 
  arrange(year) %>% 
  group_by(year) %>% 
  summarise_all(funs(mean, median, min, max))%>% 
  mutate(
    pats_val = paste0(median, "(", min, ", ", max, ")")
  )

tab4_all <- left_join(tab4, tab4_avg_pats, by = "year") %>% 
  mutate(
    ptc_rec = if_else(is.na(nrows / dplyr::lag(nrows)), "--", paste0(round(nrows / dplyr::lag(nrows) * 100, 0), "%"))
  ) %>% 
  select(year, nrows, ptc_rec, npats, median, min, max)

tab4_all$year <- as.character(tab6_all$year)

```

```{r tab4_disp, results="asis"}

kable(tab4_all, col.names = c("Year", "Total Number of Prescriptions", "Percent Change From Last Year", "Total Number of Patients", "Median number of Prescriptions per patient", "Minimum Number of Prescriptions Per Patient", "Maximum Number of Prescriptions Per Patient"), format.args = list(big.mark = ","))

```
 
```{r demographic counts, include=F}
demogCounts <- sqlQuery(sqldb, 
                  paste('SELECT gender, primary_language, race1, count(*) as nRows',

                        'FROM ',demographics,
                        'group by gender, primary_language, race1',
                        sep=" "
                        ),
                 as.is=c(TRUE,TRUE,TRUE,TRUE),
                 max=maxQryRows
                  )
demogCounts <- trimChrVars(demogCounts)

```

## Table 2. Top 5 Patient Primary Languages , Dataset: DEMOGRAPHICS 

```{r primary language, echo=F}
primLang <- demogCounts %>% 
            group_by(primary_language) %>% 
            summarise(langCnt = sum(nRows)) %>% 
            ungroup()  %>% 
            mutate(langTot = sum(langCnt), 
                  langPct = round(100*langCnt/langTot,1)
                  )

primLang <- within(merge(primLang,isoLang[c('code3B','InEnglish')],by.x='primary_language', by.y='code3B', all.x=T) %>% arrange(desc(langPct)) ,{
    InEnglish <- ifelse(!is.na(InEnglish),InEnglish, ifelse(is.na(primary_language) , 'Missing', 'Invalid'))
})

knitr::kable(primLang[1:5, c('primary_language', 'InEnglish','langCnt','langPct')], 
            col.names=     c("ISO Language","Language name ","Frequency","Percent"), 
            row.names=FALSE, 
            format.args = list(big.mark = ","))
```

## Table 3. Gender distribution , Dataset: DEMOGRAPHICS 

```{r gender , echo=F}
gender <-   demogCounts %>% 
            group_by(gender) %>% 
            summarise(count = sum(nRows)) %>% 
            ungroup()  %>% 
            mutate(totCount = sum(count), 
                  countPct = round(100*count/totCount,1)
                  )

gender <- within(merge(gender,subset(valSets,tableName=='DEMOGRAPHICS' & columnName=='GENDER')[c('code','decode')], by.x='gender', by.y='code', all.x=TRUE) %>% arrange(desc(countPct)) ,{
    decode <- ifelse(!is.na(decode),decode, ifelse(is.na(gender) , 'Missing', 'Invalid'))
})
 
knitr::kable(gender[    c('gender', 'decode','count','countPct')], 
            col.names = c("Value","Label","Frequency","Percent"), row.names=FALSE, 
            format.args = list(big.mark = ","))

```

## Table 4. Race distribution , Dataset: DEMOGRAPHICS 

```{r race , echo=F}
race <-   demogCounts %>% 
            group_by(race1) %>% 
            summarise(count = sum(nRows)) %>% 
            ungroup()  %>% 
            mutate(totCount = sum(count), 
                  countPct = round(100*count/totCount,1)
                  )

race <- within(merge(race,subset(valSets,tableName=='DEMOGRAPHICS' & columnName=='RACE1')[c('code','decode')], by.x='race1', by.y='code', all.x=TRUE) %>% arrange(desc(countPct)) ,{
    decode <- ifelse(!is.na(decode),decode, ifelse(is.na(gender) , 'Missing', 'Invalid'))
})
 
knitr::kable(race[    c('race1', 'decode','count','countPct')], 
            col.names = c("Value","Label","Frequency","Percent"), 
            row.names=FALSE, 
            format.args = list(big.mark = ","))

```

```{r encounter table, include=FALSE}

rm(diag)
gc()

# create a temp table in the SQL DB temp drive, this will not be pulled into the R workspace.
nada_ <- sqlQuery(sqldb,
                  paste('SELECT a.*, floor(DATEDIFF(day, b.birth_date, a.ADATE)/365.25) as age, b.birth_date, month(a.adate) as aMon, year(a.adate) as aYear, month(b.birth_date) as dobMon, year(b.birth_date) as dobYear',
                        'INTO #encTmp',
                        'FROM',encounters, ' as a left join (select distinct person_id, birth_date from ',demographics,') as b on a.person_id = b.person_id',
                        sep=" "
                        )
                  )

# short labels for encounter type 
encLabels <- data.frame(c('AV','ED','IP','LO','OE','RO'),
                        c('Ambulatory','ED','Inpatient','Lab only','Other','Radiology' ) ,
                        stringsAsFactors = F
)
names(encLabels)<-c('code','decode')

# overall enc summary
# read encounter table
encSumm  <- sqlQuery(sqldb,
                 paste('SELECT count(*) as rowCnt, count(distinct enc_id) as encCnt, count(distinct person_id) as personCnt, min(adate) as firstEnc, max(adate) as lastEnc',
                       'FROM #encTmp',
                       sep=" "
                       ),
                 as.is=c(TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,TRUE),
                 max=maxQryRows
                 )
encRows <- encSumm$rowCnt
encPrimKeyRows <- encSumm$encCnt
nEncPats <- encSumm$personCnt
```

#Encounter table 

Total rows in the dataset: `r encRows`

Total unique combinations of the primary key: `r encPrimKeyRows `

Total patients in the encouter table: `r nEncPats`

Month and year of first encounter: `r format(as.Date(encSumm$firstEnc),"%B %Y")`

Month and year of last encounter: `r format(as.Date(encSumm$lastEnc),"%B %Y")`

`r ifelse(encRows== encPrimKeyRows," ","The encounter table has duplicate rows, by enc_id")`


##Table 5 . Count of distinct patients by age group and year of utilization, Dataset: ENCOUNTERS

```{r echo=F}
 
# encPersonYear  <- sqlQuery(sqldb,
#                  paste('SELECT aMon, aYear, dobMon, dobYear, count(*) as personCnt',
#                        'FROM (SELECT distinct person_id, aYear, dobMon, dobYear, min(aMon) as aMon ',
#                        '      FROM #encTmp',
#                        '      GROUP BY person_id, aYear, dobMon, dobYear) as qry',
#                        'GROUP BY  aMon, aYear, dobMon, dobYear',
#                        sep=" "
#                        ),
#                  as.is=c(TRUE,TRUE,TRUE,TRUE,TRUE),
#                  max=maxQryRows
#                  )
encPersonYear  <- sqlQuery(sqldb,
                 paste('SELECT aMon, aYear, age, count(*) as personCnt',
                       'FROM (SELECT distinct person_id, aYear, min(age) as age, min(aMon) as aMon ',
                       '      FROM #encTmp',
                       '      GROUP BY person_id, aYear) as qry',
                       'GROUP BY  aMon, aYear, age',
                       sep=" "
                       ),
                 as.is=c(TRUE,TRUE,TRUE,TRUE,TRUE),
                 max=maxQryRows
                 )
encPersonAgeYear <- within(encPersonYear,{
                           adate_YM <- as.Date(paste(as.character(aYear),as.character(aMon),'01',sep='-'))
                           #dob_YM <- as.Date(paste(as.character(dobYear),as.character(dobMon),'01',sep='-'))
                           #age <- as.numeric(floor((difftime(adate_YM, dob_YM, units='days'))/365.25))
                           ageCat <- ageCatCalc(as.numeric(age))
                    }) %>% 
                    group_by(aYear, ageCat) %>%
                    summarise(Freq = sum(personCnt)) %>% 
                    group_by(aYear) %>%
                    mutate(yearTot = sum(Freq),
                           pct = round(100*Freq/yearTot,1),
                           toDisplay = paste0(Freq,' (',pct,')')
                           )  
 # transpose to square table
encPersonAgeYear_tx <- spread(encPersonAgeYear[c('ageCat','aYear','toDisplay')],aYear,toDisplay)

knitr::kable(encPersonAgeYear_tx, 
             col.names=c("Age at First Encounter in CY, N(%)",names(encPersonAgeYear_tx)[-1]), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```


Percentages are calculated with all patients within the calendar year as the denominator

## Figure 1. Count of distinct patients by age group and year of utilization, Dataset: ENCOUNTERS

```{r echo=F}
ggplot(data = encPersonAgeYear, aes(x =aYear, y = Freq, fill = ageCat)) + 
    geom_bar(stat = "identity") + labs(y='Count', x='Year', fill='Age at encounter')+theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Figure 2. Distinct patients by age group and year of utilization, as a percentage of all patients in the calendar year, Dataset: ENCOUNTERS

```{r echo=F}

ggplot(data = encPersonAgeYear, aes(x =aYear, y = pct , fill = ageCat )) + 
    geom_bar(stat = "identity") + labs(y='Percent', x='Year', fill='Age at encounter') +theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r encounter types by year and month, include=F}
encEncType_YM  <- sqlQuery(sqldb,
                 paste('SELECT aMon, aYear, encType, count(*) as rowCnt',
                       'FROM #encTmp',
                       'GROUP BY aMon, aYear, encType',
                       sep=" "
                       ),
                 as.is=c(TRUE,TRUE,TRUE,TRUE,TRUE),
                 max=maxQryRows
                 )
encEncType_YM <- within(encEncType_YM,{
               adate_YM <- as.Date(paste(as.character(aYear),as.character(aMon),'01',sep='-'))
})

```


## Table 6. Count of encounters by encounter type , Dataset: ENCOUNTERS 

```{r include=F}
encEncType <- encEncType_YM %>% group_by(encType) %>% summarize(Freq = sum(rowCnt))
encEncType <- merge(encEncType,subset(valSets,tableName=='ENCOUNTERS' & columnName=='ENCTYPE')[c('code','decode')], by.x='encType', by.y='code', all.x=TRUE) 
encEncType <- within(encEncType, {
                            pct <- round(100*Freq/sum(Freq),1)
                           # validVal <- ifelse(is.na(decode),'No','Yes')
                            decode <- ifelse(!is.na(decode),decode, ifelse(is.na(encType), 'Missing', 'Invalid'))
                            })[c('encType','decode','Freq','pct')]
```
```{r echo=FALSE}
knitr::kable(encEncType, col.names=c("Raw value","Value name",'Frequency','Percent'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))


```


##Table 7 . Count of encounters by year, Dataset: ENCOUNTERS 

```{r include=F}
encByYear <- encEncType_YM %>% group_by(aYear ) %>% summarize(Freq = sum(rowCnt))

encByYear <- within(encByYear,{
                    pct <- round(100*Freq/sum(Freq),1)
 })
encByYear$pctChng <- NA
 for(i in 2:nrow(encByYear)){
   encByYear$pctChng[i]<-
    round(100*(encByYear$Freq[i]-encByYear$Freq[i-1])/encByYear$Freq[i-1],1)
 }

encByYear$aYear <- as.character(encByYear$aYear)

```
```{r echo=FALSE}
knitr::kable(encByYear, col.names=c("Year",'Frequency','Percent','% Change from previous'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```


##Table 8. Unique patients with encounters by year, Dataset: ENCOUNTERS 

```{r include=F}
uniqPatByYear <- sqlQuery(sqldb,
                 paste('SELECT aYear, count(distinct person_id) as Freq',
                       'FROM #encTmp',
                       'GROUP BY aYear',
                       sep=" "
                       ),
                 as.is=c(TRUE,TRUE,TRUE,TRUE,TRUE),
                 max=maxQryRows
                 )

uniqPatByYear <- within(uniqPatByYear,{
                    pct <- round(100*Freq/sum(Freq),1)
 })
uniqPatByYear$pctChng<-NA
for(i in 2:nrow(uniqPatByYear)){
   uniqPatByYear$pctChng[i]<- round(100*(uniqPatByYear$Freq[i]-uniqPatByYear$Freq[i-1])/uniqPatByYear$Freq[i-1],1)
 }

uniqPatByYear$aYear <- as.character(uniqPatByYear$aYear)

```
```{r echo=FALSE}
knitr::kable(uniqPatByYear, col.names=c("Year",'Frequency','Percent','% Change from previous'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```

```{r add labels to encEncType_YM , include=F}

encTypeBymonYr <- merge(encEncType_YM,encLabels, by.x='encType', by.y='code', all.x=TRUE) 
encTypeBymonYr <- within(encTypeBymonYr,{decode <- ifelse(!is.na(decode),decode, ifelse(is.na(encType), 'Missing', 'Invalid'))})

```

##Figure 3a. Ambulatory encounters by admission year and month , Dataset: ENCOUNTERS 

```{r echo=F}
ggplot(subset(encTypeBymonYr,encType == 'AV'), aes(adate_YM, rowCnt)) + geom_line() +
  scale_x_date(date_labels = "%b %y",date_breaks='1 year') + xlab("Month of encounter") + ylab("Frequency")+theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

##Figure 3b. Other encounter types by admission year and month , Dataset: ENCOUNTERS 

```{r echo=F}
ggplot(subset(encTypeBymonYr,encType != 'AV'), aes(adate_YM, rowCnt, colour=decode)) + geom_line() +
  scale_x_date(date_labels = "%b %y",date_breaks='1 year') + xlab("Month of encounter") + ylab("Frequency")+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(color='Encounter type')

#,legend.position="bottom" "

```

 
```{r include=F}
#
# Encounter types by year
#

encTypeByYr <- encEncType_YM %>% group_by(encType, aYear ) %>% summarize(Freq = sum(rowCnt))
encTypeByYr <- merge(encTypeByYr,encLabels, by.x='encType', by.y='code', all.x=TRUE) 
encTypeByYr <- within(encTypeByYr,{decode <- ifelse(!is.na(decode),decode, ifelse(is.na(encType), 'Missing', 'Invalid'))})

```

##Figure 4a. Ambulatory encounters by admission year , Dataset: ENCOUNTERS 

```{r echo=F}
ggplot(data = subset(encTypeByYr,encType = 'AV'), aes(x =aYear, y = Freq)) + 
    geom_bar(stat = "identity") + labs(y='Count', x='Year')+theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

##Figure 4b. Other encounter types by admission year, Dataset: ENCOUNTERS  

```{r echo=F}

ggplot(data = subset(encTypeByYr,encType != 'AV'), aes(x =aYear, y = Freq, fill = decode)) + 
    geom_bar(stat = "identity") + labs(y='Count', x='Year', fill='Encounter type')+theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

##Table 9. Unique patients with encounters in the last year of data , Dataset: ENCOUNTERS

```{r include=F}


uniqPatsEncTypeLastYear <- sqlQuery(sqldb,
                 paste('SELECT encType, count(distinct person_id) as Freq',
                       'FROM #encTmp',
                       'where DATEDIFF(day, (select max(adate) from #encTmp), ADATE) between  -365 and 0',
                       'GROUP BY  encType',
                       sep=" "
                       ),
                 as.is=c(TRUE,TRUE,TRUE,TRUE,TRUE),
                 max=maxQryRows
                 )


uniqPatsEncTypeLastYearTBL <-  merge(uniqPatsEncTypeLastYear,encLabels, by.x='encType', by.y='code', all.x=TRUE)

#uniqPatsEncTypeLastYearTBL$encType <- as.character(uniqPatsEncTypeLastYearTBL$encType)
#uniqPatsEncTypeLastYearTBL <- rbind(uniqPatsEncTypeLastYearTBL, c('gh',100,NA),c('   ',200,NA))

uniqPatsEncTypeLastYearTBL <- within(uniqPatsEncTypeLastYearTBL,{
  decode <- ifelse(!is.na(decode),decode, ifelse(is.na(encType), 'Missing', 'Invalid'))
  pct <- round(100*Freq/sum(Freq),1)})


```

```{r echo=F}
knitr::kable(uniqPatsEncTypeLastYearTBL[c('encType','decode','Freq','pct')], 
             col.names=c('encType','Label','Frequency','Percent'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))
```

* Patients will be counted only once per encounter type, but also in each encounter type observed.

# Census Location

```{r census location, echo=F}
locSumm  <- sqlQuery(sqldb,
                 paste('SELECT count(*) as rowCnt, COUNT(DISTINCT person_id+CAST(loc_start AS varchar(10)) ) as primKeyCnt, count(distinct person_id) as personCnt, min(loc_start) as firstDT, max(loc_start) as lastDT',
                       'FROM  ',census_location,
                       sep=" "
                       ),
                 as.is=c(TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,TRUE),
                 max=maxQryRows
                 )

tmpLastLoc_ <- sqlQuery(sqldb,
                        paste('SELECT distinct a.*',
                              "INTO #tmpLastLoc",
                              "FROM ",census_location," AS a INNER JOIN ",
                              "(SELECT person_id, max(loc_start) as loc_start FROM",census_location,"group by person_id) AS b",
                              "ON a.person_id = b.person_id and a.loc_start=b.loc_start",
                              sep=' '
                              ),
                        max=maxQryRows
)
lastLocation <- sqlQuery(sqldb,
                         paste("SELECT stateCnty, count(*) as Freq",
                               "FROM (select distinct substring(geocode,1,5) as stateCnty, person_id FROM #tmpLastLoc) as q",
                               "GROUP BY stateCnty",
                               sep=" "),
                       as.is=c(TRUE,TRUE),
                        max=maxQryRows
                         )
lastLocation <- trimChrVars(lastLocation)
 multLastLocs <- sqlQuery(sqldb,
                          paste("SELECT count(distinct person_id) as patsMultLoc",
                                "FROM (select person_id FROM #tmpLastLoc GROUP BY person_id HAVING count(*)>1) as q",
                                sep=" "))

```

Total rows in the dataset: `r locSumm$rowCnt`

Total unique combinations of the primary key: `r locSumm$primKeyCnt `

Total patients in the location table: `r locSumm$personCnt`

Month and year of first location: `r format( as.Date(locSumm$firstDT),"%B %Y")`

Month and year of last location: `r format(as.Date(locSumm$lastDT),"%B %Y")`

`r ifelse(locSumm$rowCnt==locSumm$primKeyCnt," ","The Census Location table has duplicate rows, by person_id and loc_start")`

`r multLastLocs$patsMultLoc` patients have more than one most recent location record

##Table 10. Count of unique patients by State and county, based on most recent location, Dataset: CENSUS_LOCATION

```{r echo=F}

lastLocation <- merge(lastLocation, stateCnty[c('stateCnty','stCntyNm')], by = 'stateCnty', all.x=T) %>% arrange(desc(Freq)) %>% 
                mutate(pct = round(100*Freq/sum(Freq),1),
                       stCntyNm = ifelse(!is.na(stCntyNm), stCntyNm, ifelse( is.na(stateCnty),'Missing','Invalid'))
                     )
knitr::kable(lastLocation[c('stateCnty','stCntyNm','Freq','pct')], 
             col.names=c('State/county code','State: county','Frequency','Percent'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```

#Diagnoses

```{r echo=F}

diagSumm  <- sqlQuery(sqldb,
                 paste('SELECT count(*) as rowCnt, COUNT(DISTINCT enc_id+dx+CAST(adate AS varchar(10))+diagprovider ) as primKeyCnt, count(distinct person_id) as personCnt, min(adate) as firstDT, max(adate) as lastDT',
                       'FROM  ',diagnoses,
                       sep=" "
                       ),
                 as.is=c(TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,TRUE,TRUE),
                 max=maxQryRows
                 )


```

Total rows in the dataset: `r diagSumm$rowCnt`

Total unique combinations of the primary key: `r diagSumm$primKeyCnt `

Total patients in the table: `r diagSumm$personCnt`

Month and year of first diagnosis: `r format( as.Date(diagSumm$firstDT),"%B %Y")`

Month and year of last diagnosis: `r format( as.Date(diagSumm$lastDT),"%B %Y")`

`r ifelse( diagSumm$rowCnt==diagSumm$primKeyCnt," ","The table has duplicate rows, by enc_id dx adate diagprovider")`

##Table 11. Count of diagnoses by year , Dataset: DIAGNOSIS 

```{r echo=F}
diagsYear <- sqlQuery(sqldb,
                      paste("SELECT year(adate) as year, count(*) as Freq",
                            "FROM ",diagnoses,
                            "GROUP BY YEAR(adate)",
                            sep=" "),
                 max=maxQryRows
                 )

diagsYear <- within(diagsYear,{ pct <- round(100*Freq/sum(Freq),1)})

diagsYear$year <- as.character(diagsYear$year)

knitr::kable(diagsYear[c('year','Freq','pct')], 
             col.names=c('Year','Frequency','Percent'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```

##Table 12. Count of unique patients with at least 1 diagnoses by year , Dataset: DIAGNOSIS 

```{r echo=F}
patDiagYr <- sqlQuery(sqldb,
                      paste("SELECT year(adate) as year, count(distinct person_id) as Freq",
                            "FROM ",diagnoses,
                            "GROUP BY YEAR(adate)",
                            sep=" "),
                 max=maxQryRows
                 )
patDiagYr <- within(patDiagYr,{ pct <- round(100*Freq/sum(Freq),1)})

patDiagYr$year <- as.character(patDiagYr$year)

knitr::kable(patDiagYr[c('year','Freq','pct')], 
             col.names=c('Year','Frequency','Percent'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```

##Table 13. Average diagnoses per encounter, by year , Dataset: ENCOUNTERS 

```{r echo=F}

avgDxAll <- sqlQuery(sqldb,
                      paste("SELECT YEAR(adate) as year, count(*) as rowCnt, count(distinct enc_id) as encCnt, count(distinct person_id) as ptCnt",
                            "FROM " ,diagnoses,
                            "GROUP BY YEAR(adate)",
                            sep=" "),
                 max=maxQryRows
                 )
avgDxAll <- within(avgDxAll,{
  NDxEnc <- round(rowCnt/encCnt,1)
  NDxPat <- round(rowCnt/ptCnt,1)
})
knitr::kable(avgDxAll[c('year','rowCnt','encCnt','ptCnt','NDxEnc','NDxPat')], 
             col.names=c('Year','Diagnoses','Encounters','Patients','Dx per encounter','Dx per patient'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))


```

```{r monthly counts by enc type, echo=F}

dxByMonEncTp <- sqlQuery(sqldb,
                      paste("SELECT YEAR(adate) as year, MONTH(adate) as month, enctype, count(*) as Freq",
                            "FROM " ,diagnoses,
                            "GROUP BY YEAR(adate), MONTH(adate), enctype",
                            sep=" "),
                 max=maxQryRows
                 )

dxByMonEncTp <- within(dxByMonEncTp,{
  monYr <- as.Date(paste(as.character(year),as.character(month),'01',sep="-"))
})

dxByMonEncTp <- merge(dxByMonEncTp,encLabels, by.x='enctype', by.y='code', all.x=TRUE) 
dxByMonEncTp <- within(dxByMonEncTp,{decode <- ifelse(!is.na(decode),decode, ifelse(is.na(enctype), 'Missing', 'Invalid'))})

```

## Figure 5a. Count of diagnoses by year and month in ambulatory encounters , Dataset: DIAGNOSES 

```{r echo=F}

ggplot(subset(dxByMonEncTp,enctype == 'AV'), aes(monYr, Freq)) + geom_line() +
  scale_x_date(date_labels = "%b %y",date_breaks='1 year') + xlab("Month of encounter") + ylab("Frequency")+theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Figure 5b. Count of diagnoses by year and month, other encounter types , Dataset: DIAGNOSES 

```{r echo=F}
ggplot(subset(dxByMonEncTp,enctype != 'AV'), aes(monYr, Freq, colour=decode)) + geom_line() +
  scale_x_date(date_labels = "%b %y",date_breaks='1 year') + xlab("Month of encounter") + ylab("Frequency")+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(color='Encounter type')

#,legend.position="bottom" "

```

##Table 14. Top 10 diagnoses by diagnosis code and codetype , Dataset: DIAGNOSES

```{r echo=F}
top10Dx <- sqlQuery(sqldb,
                      paste("SELECT dx, dx_codetype, count(*) as Freq",
                            "FROM " ,diagnoses,
                            "GROUP BY dx, dx_codetype",
                            sep=" "),
                 max=maxQryRows,
                 as.is=c(TRUE,TRUE,TRUE )
                 )

top10Dx <- within(top10Dx, {
  pct <- round(100*Freq/sum(Freq), 1)
  dxCdTp  <- paste(dx_codetype,'-',gsub('.','',dx, fixed=T), sep='')
  }) %>% arrange(desc(Freq))

top10Dx <- top10Dx[1:10,]
top10Dx <- trimChrVars(top10Dx)
top10Dx <- merge(top10Dx, icdLU, by.x = 'dxCdTp', by.y = 'dxCdTp', all.x=T)
top10Dx <- within(top10Dx,{
                  dxDesc <- ifelse(!is.na(dxDesc),dxDesc, ifelse(is.na(dx), 'Missing', 'Invalid'))
                  
})

knitr::kable(top10Dx[c('dx_codetype','dx','dxDesc','Freq','pct')], 
             col.names=c('Code Type','Diagnoses','Description','Frequency','Percent'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```

##Table 15. Average diagnoses per encounter, by encounter type, Dataset: DIAGNOSES

```{r echo=F}

NAvgDxEnc <- sqlQuery(sqldb,
                      paste(
                     'SELECT enctype, avg(Freq) as NAvgDx',
                     'FROM (',
                       'SELECT enc_id, enctype, count(*)*1.0 as Freq',
                       'FROM ', diagnoses,
                       'GROUP BY enc_id, enctype) as q',
                     'GROUP BY enctype',
                     sep=" "
                      ),
                 max=maxQryRows,
                 as.is=c(TRUE )
                     )

NAvgDxEnc <- within(
               merge(NAvgDxEnc,subset(valSets,tableName=='DIAGNOSES' & columnName=='ENCTYPE')[c('code','decode')], by.x='enctype', by.y='code', all.x=TRUE),{
                 NAvgDx <- round(as.numeric(NAvgDx),1)
                 decode <- ifelse(!is.na(decode),decode, ifelse(is.na(enctype) , 'Missing', 'Invalid'))

                 
               })

knitr::kable(NAvgDxEnc[c('enctype','decode','NAvgDx')], 
             col.names=c('Encounter Type','Label','Average diagnoses'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```


#VITAL_SIGNS

```{r echo=F}

vitalSumm  <- sqlQuery(sqldb,
                 paste('SELECT count(*) as rowCnt, COUNT(DISTINCT person_id+CAST(measure_date AS varchar(10))+CAST(measure_time AS varchar(20)) ) as primKeyCnt, count(distinct person_id) as personCnt, min(measure_date) as firstDT, max(measure_date) as lastDT',
                       'FROM  ',vital_signs,
                       sep=" "
                       ),
                 max=maxQryRows
                 )


```

Total rows in the dataset: `r vitalSumm$rowCnt`

Total unique combinations of the primary key: `r vitalSumm$primKeyCnt `

Total patients in the table: `r vitalSumm$personCnt`

Month and year of first record: `r format(as.Date(vitalSumm$firstDT),"%B %Y")`

Month and year of last record: `r format(as.Date(vitalSumm$lastDT),"%B %Y")`

`r ifelse(vitalSumm$rowCnt==vitalSumm$primKeyCnt," ","The table has duplicate rows, by person_id, measure_date, measure_time")`


##Table 16. Vital Sign Summary, Dataset: VITAL_SIGNS
```{r echo=F}

# be sure to include space breaks in any re-design
vitalVarSumm <- function(var,label){
  
  sqlQuery(sqldb,
                 paste('SELECT variable, minimum, q1, median, q3, maximum, count(*) as nobs, count(distinct person_id) as npats ',
                       'FROM (',
                       "SELECT dummy, variable, person_id, ",var,
                       ', percentile_cont(0.0) WITHIN GROUP (ORDER BY ',var,') OVER(PARTITION BY dummy) as  Minimum',
                       ', percentile_cont(0.25) WITHIN GROUP (ORDER BY ',var,') OVER(PARTITION BY dummy) as  q1',
                       ', percentile_cont(0.5) WITHIN GROUP (ORDER BY ',var,') OVER(PARTITION BY dummy) as  Median',
                       ', percentile_cont(0.75) WITHIN GROUP (ORDER BY ',var,') OVER(PARTITION BY dummy) as  q3',
                       ', percentile_cont(1.0) WITHIN GROUP (ORDER BY ',var,') OVER(PARTITION BY dummy) as  maximum ',
                       'FROM  ',
                         "(SELECT 1 as dummy, '",label,"' as variable, * FROM ",vital_signs,") as q ",
                       'WHERE ',var,' > 0 ) as q2 ',
                       'GROUP BY variable, minimum, q1, median, q3, maximum ',
                       sep=""
                       ),
                 max=maxQryRows,
                 as.is=c(TRUE)
                 )

}
vlst <- do.call('rbind',
                list(vitalVarSumm('ht','Height'),
                     vitalVarSumm('wt','Weight'),
                     vitalVarSumm('systolic','Systolic BP'),
                     vitalVarSumm('diastolic','Diastolic BP')))

vlst <- within(vlst,{
  minimum <- round(as.numeric(minimum),1)
  q1      <- round(as.numeric(q1),1)
  median  <- round(as.numeric(median),1)
  q3      <- round(as.numeric(q3),1)
  maximum <- round(as.numeric(maximum),1)
})
knitr::kable(vlst[c('variable','nobs','npats','minimum','q1','median','q3', 'maximum')], 
             col.names=c('Vital record','Non-null obs.','Unique patients with non-null obs','Minimum','Q1','Median','Q3', 'Maximum'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))

```

##Table 17. Row counts of vital records by year , Dataset: VITAL_SIGNS

```{r echo=F}

vitalsByYr <- sqlQuery(sqldb,
                 paste('SELECT year(measure_date) as year, count(*) as Freq',
                       'FROM  ',vital_signs,
                       'GROUP BY year(measure_date)',
                       sep=" "
                       ),
                 max=maxQryRows
                 )

vitalsByYr <- within(vitalsByYr,{
  pct <- round(100*Freq/sum(Freq), 1)
})

vitalsByYr$year <- as.character(vitalsByYr$year)

knitr::kable(vitalsByYr[c('year','Freq','pct')], 
             col.names=c('Year','Frequency','Percent'), 
             row.names=FALSE, 
             format.args = list(big.mark = ","))


```

##Figure 6. Count of vital signs by year and month, Dataset: VITAL_SIGNS

```{r echo=F}

vitalsByMonYr <- sqlQuery(sqldb,
                 paste('SELECT year(measure_date) as year, month(measure_date) as month',
                       ', sum(case when ht>0 then 1 else 0 end) as height',
                       ', sum(case when wt>0 then 1 else 0 end) as weight',
                       ', sum(case when systolic>0 then 1 else 0 end) as systolic',
                       ', sum(case when diastolic>0 then 1 else 0 end) as diastolic',
                       'FROM  ',vital_signs,
                       'GROUP BY year(measure_date), month(measure_date)',
                       sep=" "
                       ),
                 max=maxQryRows
                 )

vitalsByMonYr <- within(vitalsByMonYr,{
  monYr <- as.Date(paste(as.character(year),as.character(month),'01', sep='-'))
}) %>% subset(select=-c(month,year))

vitalsByMonYrTx <- gather(vitalsByMonYr, 'Vital_record','Count', 1:4)

ggplot(vitalsByMonYrTx, aes(monYr, Count, colour=Vital_record)) + geom_line() +
  scale_x_date(date_labels = "%b %y",date_breaks='1 year') + xlab("Month of encounter") + ylab("Frequency")+theme(axis.text.x = element_text(angle = 45, hjust = 1))+labs(color='Vital type')

```

####Total program run time:
```{r Calculate program run time, echo=F}
# close odbc
 close(sqldb)

endTime <- Sys.time()
endTime - startTime 
```
